---
title: "Pacific Giant Salamanders"
author: "Allison Bailey"
date: "11/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Attach required packages
library(tidyverse)
library(janitor)
library(here)
library(hrbrthemes)
library(viridis)
library(ggthemes)
library(paletteer)
library(DT)
library(plotly)
library(expss)
library(kableExtra)
library(gmodels)
library(PairedData)
library(car)
library(ggbeeswarm)

```

#### **Introduction**


```{r, include=FALSE}
# Read in csv titled "mack_creek_vertebrates.csv"

mack_data <- read_csv("mack_creek_vertebrates.csv")

mack_clean <- mack_data %>%
  clean_names()
```

```{r, include=FALSE}
mack_date <- mack_clean %>% 
  mutate(date_new = lubridate::mdy(sampledate)) %>%
  mutate(day_new = lubridate::day(date_new)) %>%
  mutate(month_new = lubridate::month(date_new)) %>%
  dplyr::select(-sampledate) %>%
  filter(species == "DITE") %>%
  filter(unittype == "P" | unittype == "C" | unittype == "SC")
```


```{r, include=FALSE}
# Notes on Effect Size (Cohen) prior to conducting data analysis:
# 1. The effect size is the main finding of a quantitative study. While a P value can inform the reader whether an effect exists, the P value will not reveal the size of the effect. In reporting and interpreting studies, both the substantive significance (effect size) and statistical significance (P value) are essential results to be reported.

# 2. Effect sizes should be reported in a paper's Abstract and Results sections.

# you must determine what number of subjects in the study will be sufficient to ensure (to a particular degree of certainty) that the study has acceptable power to support the null hypothesis. That is, if no difference is found between the groups, then this is a true finding

# 3. The indices fall into two main study categories, those looking at effect sizes between groups and those looking at measures of association between variables (table 1). For two independent groups, effect size can be measured by the standardized difference between two means, or mean (group 1) – mean (group 2) / standard deviation.

# 4. Cohen classified effect sizes as small (d  =  0.2), medium (d  =  0.5), and large (d ≥ 0.8).

# Between group means, the effect size can also be understood as the average percentile distribution of group 1 vs. that of group 2 or the amount of overlap between the distributions of interventions 1 and 2 for the two groups under comparison. For an effect size of 0, the mean of group 2 is at the 50th percentile of group 1, and the distributions overlap completely (100%)—that is , there is no difference. 
```

```{r, include=FALSE}

# Notes on differences in MEANS (ANOVA) prior to conducting data analysis:
# 1. Null Hypothesis (H0): All means are equal (in other words, all groups are from populations with the same mean). Alternative Hypothesis (HA): At least two group means are NOT equal (that means that just two could be different, or they could ALL be different)

# 2. One-Way ANOVA (also called single-factor ANOVA) There is a single factor (variable), with at least 3 levels, where we are trying to compare means across the different levels of that factor.

# Variance test: Levene’s Test (most common). Generally, if the largest sample variance is < 4× the smallest sample variance, the variances are close enough

# 3. Make sure to use a Density/Histogram plot to visually see the differences in variances.

# 4. If the distance between groups is large (high between groups variance) compared to the distance between observations within groups (low within groups variance), then we might have enough evidence to conclude that samples are drawn from populations with different means.

# if you decide that at least two means differ significantly, use the Tukey’s test for Honestly Significant Differences (Tukey’s HSD)

```

```{r include=FALSE}
# Notes on Chi-Squared
# Compare proportions.  Good for "counts"
# Null hypothesis: No association between variables studied. The null hypothesis is independence. Alternative hypothesis: There is an association between variables studied (non-independence). 
# Cells contain frequencies (counts)
# You should do a Chi-Square Test if:
# - Levels are mutually exclusive (an outcome can contribute to only one count)
# - Each subject contributes only one value to dataset (e.g. no longitudinal data with same subjects)
# - Groups are independent (e.g. no paired groups)
# - wo categorical variables
# - Value of the cell expecteds should be 5 or more in at least 80% of the cells, and no cell should have an expected of less than one. This assumption is most likely to be met if the sample size equals at least the number of cells multiplied by 5. 

```

```{r include=FALSE}
# Notes on Two Sample T-Test:
# Two-Sample t-test:

# - To compare means of two samples

# - Null hypothesis: no significant difference (difference in means = 0)

# - t-statistic: based on difference in sample means, sd, n

# - If the difference in means is very unlikely to occur by random chance if samples are drawn from identical populations, then we may decide that they ARE significantly different 

```


```{r echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 1.** Graph visualizes the comparison of annual Salamander counts in Old Growth and Clear Cut sections of Mack Creek.*"}
# Results A: Visually compare annual salamander counts in old growth (OG) and clear cut (CC) sections of Mack Creek. For all years when Pacific giant salamander observations were recorded in the study, find the total annual Pacific giant salamander counts for the two sections of Mack Creek. Create a finalized graph to show the changes in annual counts for each section over time, add a figure caption, and briefly describe the trends in text. 

sal_abundance <- mack_date %>%
    dplyr::select(year, section, species) %>%
  group_by(year, section, species) %>%
  tally()

sal_total <- sal_abundance %>%
  group_by(section, species, year) %>%
  summarise(
    total = sum(n)
  )

sal_total_plot <- ggplot(sal_total, aes(x=year, y=total)) +
    geom_line(aes(color=section), size=1.2 ) +
    theme_ipsum() +
    theme(
      legend.position="right",
      plot.title = element_text(size=14)
    ) +
    ggtitle("Total Salamanders Counted at Mack Creek") +
   scale_color_viridis(discrete=TRUE, option="cividis") +
  labs(x = "Year",
       y = "Salamander Count")

sal_total_plot
```


```{r echo=FALSE, include=FALSE, warning=FALSE}
# Results B: Table of 2017 salamander counts by channel classification (pool, cascades and side-channel) in old growth (OG) and clear cut (CC) sections of Mack Creek. Using only Pacific giant salamander observations from 2017, create a finalized table showing the counts and proportions of salamanders observed in different channel classifications (pool, cascade, or side-channel) within Mack Creek for the two sections (OG and CC). Add a table caption above the table. Note: We’re only interested in Pacific giant salamanders observed in the creek, so you should exclude salamanders observed in isolated pools (IP) disconnected from the channel. 

sal_summary <- mack_date %>%
  dplyr::select(unittype, section, year, species) %>%
  filter(unittype == "P" | unittype == "C" | unittype == "SC") %>%
  filter(year == 2017) %>%
  dplyr::select(-year) %>%
  group_by(unittype, section) %>%
  tally()


kable(sal_summary, col.names = c("Channel Classification",
                           "Section",
                           "Total Count")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))

```



```{r, include=FALSE}
# Results C: Using the 2017 counts found in Results B above, answer: is there a significant difference in where in the channel Pacific giant salamanders are located (pool, cascade or side channel) between the two sections (old growth and clear cut)? Another way to ask this: Regarding the locations where salamanders were observed in Mack Creek in 2017, is there a significant effect of forest condition (old growth or clear cut) on where in the channel salamanders are found (channel classification)? Report your statistical findings in text, in the context of the actual counts / proportions you present in the table. 

sal_count_date <- mack_date %>%
  dplyr::select(unittype, section, year, date_new) %>%
  filter(unittype == "P" | unittype == "C" | unittype == "SC") %>%
  filter(year == 2017) %>%
  group_by(date_new, unittype, section) %>%
  tally()
  

```



```{r echo=FALSE, include=FALSE, warning=FALSE}

#  fig.cap = "***Figure 1.**Table of 2017 salamander counts by channel classification (pool, cascades and side-channel) in old growth (OG) and clear cut (CC) sections of Mack Creek.*"}
# First, filter then get a table of counts & proportions: 


sal_count_table <- sal_summary %>% 
  pivot_wider(names_from = unittype, values_from = n)
 
sal_count_table
```


```{r echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 3.** Table shows the number and proportion of counts per channel in relation to old growth and clear cut sections*"}
sal_adorn <- sal_count_table %>% 
  janitor::adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")

kable(sal_adorn) %>% 
  kable_styling()
```
### 8. Chi-Square test for independence

Chi-square is a test for associations between two or more nominal (categorical variables), usually comparing counts within groups across levels. 

What we're really asking: are the proportions within different levels in a group so different *across* multiple groups that we actually think there's an effect of *group*? 

**Null hypothesis**: There is no significant association between old growth and clear cut sections (or, old growth and clear cut are independent)

**Alternative hypothesis**: There IS a significant association between old growth and clear cut (or, old growth and clear cut are NOT independent)

```{r, include=FALSE}
# First, get just a contingency table of counts we just want the six count values in a df, nothing else: 

chi_counts <- sal_count_table %>% 
  dplyr::select(-section)

chi_counts

```

```{r, include=FALSE}

# B. Run a chi-square test for independence using `chisq.test()`:

my_channel_chi <- chisq.test(chi_counts)

my_channel_chi

summary(my_channel_chi)

```

There is a *not* a significant assocation between old growth and clear cut sections ($\chi$^2^(`r my_channel_chi$parameter`)) = `r my_channel_chi$statistic` and ($p-value$(`r my_channel_chi$parameter`)) = `r my_channel_chi$p.value` Therefore, the **Null hypothesis** stating 'There is no significant association between old growth and clear cut sections (or, old growth and clear cut are independent)' remains true.

### Welch two sample ttest for comparing difference in means of clear cut and old growth sections.


**Null hypothesis**: There is no significant difference in weight mean between old growth and clear cut sections (or, old growth and clear cut weight means not significantly different)

**Alternative hypothesis**: There IS a significant difference in weight mean between old growth and clear cut (or, old growth and clear cut are significantly differenct)


```{r, include=FALSE}
# Results D. Compare weights of Pacific giant salamanders in clear cut and old growth forest sections of the creek in 2017. Only considering creek section (OG or CC) as a variable (not further factoring by channel classification), answer: Is there a significant difference in mean weights for Pacific giant salamanders observed in the two forest sections (clear cut and old growth) in 2017? 

sal_stats <- mack_date %>% 
  filter(year == 2017) %>%
  group_by(section) %>% 
  summarize(
    mean_weight = mean(weight, na.rm = TRUE),
    sd_weight = sd(weight, na.rm = TRUE),
    sample_size = n(),
    se_weight = sd(weight, na.rm = TRUE) / sqrt(n()),
    var_weight = var(weight, na.rm = TRUE)
  )

sal_stats

kable(sal_stats, col.names = c("Section",
                           "Mean Weight",
                           "SD Weight", "Sample Size",
                           "SE Weight",
                           "Var Weight"),
        caption = 
            "Figure 5. The table below includes the means, standard deviations, and sample weights for Clear Cut and Old Growth sections in 2017.") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))


# lRun a two-sample T-Test to see if there is a significant difference in the means of clear cut vs. old growth sections. 

# Question: Are clear cut salamander significantly different in weight than old growth slamanders(using a significance level of 0.05)?

sec_welch <- mack_date %>%
  dplyr::select(section, year, weight, unittype) %>%
  filter(unittype == "P" | unittype == "C" | unittype == "SC") %>%
  filter(year == 2017) %>%
  dplyr::select(-year) %>%
  group_by(weight, section)

cc_sample <- sec_welch %>%
  filter(section == "CC") %>%
  pull(weight)

t.test(cc_sample)

og_sample <- sec_welch %>%
  filter(section == "OG") %>%
  pull(weight)

t.test(og_sample)

section_ttest <- t.test(cc_sample,og_sample)

section_ttest

# A resulting p-value under 0.05 means that variances are not equal and than further parametric tests such as ANOVA are not suited. We should use chi-squared.
```

```{r}
# Run a density plot to visualize if there's a significant difference in means. 

section_density <- sec_welch

sec_affil <- ggplot(section_density, aes(x=weight, group = section)) +
    geom_density(aes(color = section, fill = section), alpha=0.8) +
    ggtitle("Weight of Salamanders at Clear Cut vs. Old Growth Sections") +
  theme_ipsum() +
  scale_color_viridis(discrete = TRUE, option = "cividis", guide = FALSE) +
  scale_fill_viridis(discrete = TRUE, option = "cividis", name = "Section") +
  geom_vline(xintercept = 3.7, color = "darkgrey", size = 0.5, linetype = 2)

sec_affil +
   theme(
    plot.title=element_text( hjust=0.5, vjust=0.5, face='bold')
    )

```


“Weight of Salamanders located in Clear Cut seciton (7.77587 ± 9.9 g [mean ± sd],  n = 368) and Old Growth section (6.701818 ± 9.035g [mean ± sd], n = 320) are *not* significantly different (t(683.21) = 1.4859, p = 0.1378).”


```{r echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 3.** Graph visualizes the relativity of observed weights in relation to the mean in salamander weights in old growth and clear cut sections"}
# Results C continued.

sal_weight <- mack_date %>%
  dplyr::select(year, section, weight) %>%
  filter(year == 2017)

sal_weight_plot <- ggplot() +
  geom_beeswarm(data = sal_weight, 
                aes(x = section, 
                    y = weight,
                    ),
                size = 3,
                alpha = 0.6,
                pch = 16,
                color = "gray70") +
  scale_x_discrete(labels = c("Clear Cut", "Old Growth")) +
  geom_point(data = sal_stats, aes(x = section, y = mean_weight), 
             color = "firebrick",
             size = 4) +
  geom_errorbar(data = sal_stats, 
                aes(x = section, 
                    ymin = mean_weight - sd_weight,
                    ymax = mean_weight + sd_weight),
                color = "firebrick",
                width = 0.1) +
  theme_light()

sal_weight_plot

```

<b/r>
In the graphic above Clear Cut and Old Growth salamander weights are reported. The bar heights represent their mean value. The red error bar gives information on how the individual observations are dispersed around the average. It appears that measurements in Old Growth section are slightly more precise than in Clear Cut.  


```{r echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 3.** Graph visualizes the skewed distribution in salamander weights in old growth and clear cut sections"}
# We need to see if the weight in Salamanders is normally distrubted between the two sections. To do this. We will use a qq plot.

ggplot(data = mack_date, aes(sample = weight)) +
  geom_qq(aes(color = section),
                 alpha = 0.5,
                 show.legend = FALSE
          ) +
  facet_wrap(~section, scales = "free") +
  scale_color_viridis(discrete = TRUE, option = "cividis")

# Based on the results of our qq plot, the data is NOT normally distributed.
```

Based on the results of our qq plot, the data is NOT normally distributed in either section. The weight is not normally distributed therefore, utilizing the mean in our report may have skewed our data.  Comparing Medians and running non-parametric Wilcox Test and visualizing the data through a boxplot would've been a safer statistical analysis. 

```{r echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 8.** Table shows median weights for each channel"}
weight_channel <- mack_date %>%
  dplyr::select(unittype, weight, year) %>%
  filter(year == 2017) %>%
  filter(unittype == "P" | unittype == "C" | unittype == "SC")

weight_summary <- weight_channel %>%
  filter(year == 2017) %>%
  group_by(unittype) %>% 
  summarize(
    mean_weight = mean(weight, na.rm = TRUE),
    sd_weight = sd(weight, na.rm = TRUE),
    sample_size = n(),
    se_weight = sd(weight, na.rm = TRUE) / sqrt(n()),
    var_weight = var(weight, na.rm = TRUE)
  )

kable(weight_summary) %>% 
  kable_styling()

```


```{r, echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 9.** Plot visualizes the mean salamander weights in the different channels"}
channel_weight_plot <- ggplot() +
  geom_beeswarm(data = weight_channel, 
                aes(x = unittype, 
                    y = weight,
                    ),
                size = 3,
                alpha = 0.6,
                pch = 16,
                color = "gray70") +
  scale_x_discrete(labels = c("C", "P", "SC")) +
  geom_point(data = weight_summary, aes(x = unittype, y = mean_weight), 
             color = "firebrick",
             size = 4) +
  geom_errorbar(data = weight_summary, 
                aes(x = unittype, 
                    ymin = mean_weight - sd_weight,
                    ymax = mean_weight + sd_weight),
                color = "firebrick",
                width = 0.1) +
  theme_light()

channel_weight_plot

```

```{r, echo=FALSE, include=FALSE}
channel_aov <- aov(weight ~ unittype, data = weight_channel)
summary(channel_aov)
```
We retain the alternative hypothesis (mean weights for the 3 channels are NOT the same). But are they all different from each other? Or just two? 

```{r, echo=FALSE, include=FALSE}
# comparing to Type III ANOVA

channel_lm <- lm(weight ~ unittype, data = weight_channel)
channel_aov_t3 <- Anova(channel_lm)
channel_aov_t3

# Now we know about data structure for one-way ANOVAs, but we also need to make sure we satisfy the assumptions:

```

```{r echo=FALSE, include=FALSE}
# Normality
# Samples are independent
# Population variances are equal

# **ANOVA is relatively robust to violations of the normality assumption and the equal variance assumption. No, that doesn’t mean you can just use an ANOVA without looking at your data.

# Variance test: Levene’s Test (most common)

leveneTest(weight ~ unittype, data = weight_channel) # Not significant (we retain the null hypothesis of equal variances)


```

```{r echo=FALSE, include=FALSE}
TukeyHSD(channel_aov)

# Conclusion: means for ALL pairwise combinations are significantly different
```

```{r echo=FALSE, include=FALSE}
# Now I will find just how big of a difference SC-P really have by running a Cohen.d Test.

pool_sample <- weight_channel %>%
  filter(unittype == "P") %>%
  pull(weight)

t.test(pool_sample)

side_sample <- weight_channel %>%
  filter(unittype == "SC") %>%
  pull(weight)

t.test(side_sample)

sc_p_ttest <- t.test(pool_sample, side_sample)

sc_p_ttest

channel_cohen <- effsize::cohen.d(pool_sample, side_sample, na.rm = TRUE)

channel_cohen

summary(channel_cohen)
```

Conclusion: Mean weights did *not* differ significantly between Pool (9.29 +/- 1.56) and Cascades (7.52 +/- 0.43) nor the Cascades and Side Channel (5.68 +/- 0.65).  However, there is a significant difference between Side Channel and Pool (one-way ANOVA with post-hoc Tukey's HSD, F(2) = 4.2156, SC-P *p adj* < 0.05). However, after running the difference against a Cohen's d test and receiving a value of approximately 0.352 it can be concluded that the difference between Pool and Side channels is small.


```{r echo=FALSE, include=TRUE, warning=FALSE, fig.cap="***Figure 3.** Graph visualizes the skewed distribution in salamander weights in the different channels"}



# We need to see if the weight in Salamanders is normally distrubted between the two sections. To do this. .

ggplot(data = mack_date, aes(sample = weight)) +
  geom_qq(aes(color = unittype),
                 alpha = 0.5,
                 show.legend = FALSE
          ) +
  facet_wrap(~unittype, scales = "free") +
  scale_color_viridis(discrete = TRUE, option = "cividis")

# Based on the results of our qq plot, the data is NOT normally distributed.
```

The weight is not normally distributed therefore, utilizing the mean in our report may have skewed our data.  Comparing Medians and running non-parametric Wilcox Test and visualizing the data through a boxplot would've been a safer statistical analysis. 